### 2023-08-08 proofs/tex/norm-lemmas.tex
wip
Transitivity of ≤
wip: leftovers

\begin{lemma}[Soundness of quantifier normalization]
  \label{lemma:normalization-soundness}
  \hfill

  \begin{itemize}
    \item[$-$] $[[iN ≈ nf(iN)]]$
    \item[$+$] $[[iP ≈ nf(iP)]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Mutual induction on $[[nf(iN) = iM]]$ and $[[nf(iP) = iQ]]$.
  Let us consider how this judgment is formed:
  \begin{caseof}
    \item{\nameref{\ottdruleNrmNVarLabel} and \nameref{\ottdruleNrmPVarLabel}}\\ By
      the corresponding equivalence rules.
    \item{\nameref{\ottdruleNrmShiftULabel}, \nameref{\ottdruleNrmShiftDLabel},
        and \nameref{\ottdruleNrmArrowLabel}}\\
      By the induction hypothesis and the corresponding congruent equivalence rules.
    \item{\nameref{\ottdruleNrmForallLabel}}, i.e. $[[nf(∀pas.uN) = ∀pas'.uN']]$ \label{case:norm-soundness:forall}\\
      From the induction hypothesis, we
      know that $[[iN ≈ iN']]$. In particular, by \cref{lemma:equiv-fv}, $[[fv
        iN]] \equiv [[fv iN']]$.
      Then by \cref{lemma:ord-soundness}, $[[{pas'}]]
      \equiv [[{pas} ∩ fv iN']] \equiv [[{pas} ∩ fv iN]]$, and thus,
      $[[{pas'} ∩ fv iN']] \equiv [[{pas} ∩ fv iN]]$.

@@

@@

\begin{lemma}[Distributivity of normalization over substitution]
  \label{lemma:norm-subst-distr} Normalization of a term distributes over substitution.
  Suppose that $[[Γ2 ⊢ σ : Γ1]]$, i.e. $[[σ]]$ maps variables from $[[Γ1]]$ into types
  taking free variables from $[[Γ2]]$.
  Then
  \begin{itemize}
  \item[$-$] $[[nf([σ]iN)]] = [[ [nf(σ)] nf(iN) ]]$
  \item[$+$] $[[nf([σ]iP)]] = [[ [nf(σ)] nf(iP) ]]$
  \end{itemize}
  where $[[nf(σ)]]$ means pointwise normalization: $[[ [nf(σ)] α⁻]] = [[nf([σ]
  α⁻)]]$. %, and .
\end{lemma}
\begin{proof}
  Mutual induction on $[[iN]]$ and $[[iP]]$.
  \begin{caseof}
    \item $[[iN]]$ = $[[na]]$ \\
      \label{case:norm-subst-distr-var}
      $[[nf([σ]iN)]] = [[ nf([σ]na) ]] = [[ [nf(σ)]na ]] $.

>>>>

\begin{lemma}[Soundness of normalization]
  \label{lemma:normalization-soundness}
  \hfill
  \begin{itemize}
    \item[$-$] $[[iN ≈ nf(iN)]]$
    \item[$+$] $[[iP ≈ nf(iP)]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Mutual induction on $[[nf(iN) = iM]]$ and $[[nf(iP) = iQ]]$.
  Let us consider how this judgment is formed:
  \begin{caseof}
    \item{\nameref{\ottdruleNrmNVarLabel} and \nameref{\ottdruleNrmPVarLabel}}\\ By
      the corresponding equivalence rules.
    \item{\nameref{\ottdruleNrmShiftULabel}, \nameref{\ottdruleNrmShiftDLabel},
        and \nameref{\ottdruleNrmArrowLabel}}\\
      By the induction hypothesis and the corresponding congruent equivalence rules.
    \item{\nameref{\ottdruleNrmForallLabel}}, i.e. $[[nf(∀pas.uN) = ∀pas'.uN']]$ \label{case:norm-soundness:forall}\\
      From the induction hypothesis, we
      know that $[[iN ≈ iN']]$. In particular, by \cref{lemma:equiv-fv}, $[[fv
        iN]] \equiv [[fv iN']]$.
      Then by \cref{lemma:ord-soundness}, $[[{pas'}]]
      \equiv [[{pas} ∩ fv iN']] \equiv [[{pas} ∩ fv iN]]$, and thus,
      $[[{pas'} ∩ fv iN']] \equiv [[{pas} ∩ fv iN]]$.

@@

\begin{lemma}[Soundness of normalization of algorithmic types]
  \label{lemma:normalization-soundness-alg}
  \hfill
  \begin{itemize}
    \item[$-$] $[[uN ≈ nf(uN)]]$
    \item[$+$] $[[uP ≈ nf(uP)]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  The proof coincides with the proof of \cref{lemma:normalization-soundness}.
\end{proof}

@@

\begin{lemma}[Distributivity of normalization over substitution]
  \label{lemma:norm-subst-distr} Normalization of a term distributes over substitution.
  Suppose that $[[σ]]$ is a substitution, $[[iN]]$ and $[[iP]]$ are types. Then
    \begin{itemize}
      \item[$-$] $[[nf([σ]iN)]] = [[ [nf(σ)] nf(iN) ]]$
      \item[$+$] $[[nf([σ]iP)]] = [[ [nf(σ)] nf(iP) ]]$
  \end{itemize}
  where $[[nf(σ)]]$ means pointwise normalization: $[[ [nf(σ)] α⁻]] = [[nf([σ]α⁻)]]$.
\end{lemma}
\begin{proof}
  Mutual induction on $[[iN]]$ and $[[iP]]$.
  \begin{caseof}
    \item $[[iN]]$ = $[[na]]$ \\
      \label{case:norm-subst-distr-var}
      $[[nf([σ]iN)]] = [[ nf([σ]na) ]] = [[ [nf(σ)]na ]] $.

### 2023-08-08 proofs/tex/ord-lemmas.tex
Prove the UB lemma
wip: leftovers

\begin{lemma}[Completeness of variable ordering]
  \label{lemma:ord-completeness}
  Variable ordering is invariant under equivalence.

  \begin{itemize}
  \item[$-$] For $[[iN ≈ iM]]$ and any $[[varset]]$,
    $[[ord varset in uN]] = [[ord varset in uM]]$ (as lists)

  \item[$+$] For $[[iP ≈ iQ]]$ and any $[[varset]]$,
    $[[ord varset in uP]] = [[ord varset in uQ]]$ (as lists)
  \end{itemize}
\end{lemma}
\begin{proof}
  Mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$.
\end{proof}

>>>>

\begin{lemma}[Ordering is not affected by independent substitutions]
  \label{lemma:ord-sigma}
  Suppose that $[[Γ2 ⊢ σ : Γ1]]$, i.e. $[[σ]]$ maps variables from $[[Γ1]]$ into types
  taking free variables from $[[Γ2]]$, and $[[varset]]$ is a set of variables
  disjoint with both $[[Γ1]]$ and $[[Γ2]]$, 
  $[[iN]]$ and $[[iP]]$ are types. Then
    \begin{itemize}
  \item[$-$] $[[ ord varset in [σ]iN ]] = [[ord varset in iN ]]$
  \item[$+$] $[[ ord varset in [σ]iP ]] = [[ord varset in iP ]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  Mutual induction on $[[iN]]$ and $[[iP]]$.
  \begin{caseof}
    \item $[[iN = na]]$ \\
      If $[[na ∉ {Γ1}]]$ then $[[ [σ]na = na ]]$ and $[[ ord varset in [σ]na ]] = [[ ord varset in na ]]$, 
      as requried.
      If $[[na ∊ {Γ1}]]$ then $[[na ∉ varset]]$, so $[[ ord varset in na ]] = [[·]]$.
      Moreover, $[[Γ2 ⊢ σ : Γ1]]$ means $[[ fv([σ]na) ⊆ {Γ2} ]]$, and thus, 
      as a set, $[[ ord varset in [σ]na ]] = [[varset ∩ fv([σ]na)]] \subseteq [[varset ∩ {Γ2}]] = [[·]]$.
    \item $[[iN = ∀pas.iM]]$\\
      We can assume $[[{pas} ∩ {Γ1} = ∅]]$
      and $[[{pas} ∩ varset = ∅]]$. Then

      $\begin{aligned}[t]
         [[ ord varset in [σ]iN ]] &= [[ ord varset in [σ]∀pas.iM ]] \\
                                   &= [[ ord varset in ∀pas.[σ]iM ]]\\
                                   &= [[ ord varset in [σ]iM ]]
                                   && \text{by the induction hypothesis}\\
                                   &= [[ ord varset in iM ]]\\
                                   &= [[ ord varset in ∀pas.iM ]]\\
                                   &= [[ ord varset in iN ]]
       \end{aligned}$
    \item $[[iN = ↑iP]]$\\
       $\begin{aligned}[t]
        [[ ord varset in [σ]iN ]] &= [[ ord varset in [σ]↑iP ]] \\
                                   &= [[ ord varset in ↑[σ]iP ]]
                                   && \text{by the definition of substitution}\\
                                   &= [[ ord varset in [σ]iP ]]
                                   && \text{by the induction hypothesis}\\
                                   &= [[ ord varset in iP ]]
                                   && \text{by the definition of substitution}\\
                                   &= [[ ord varset in ↑iP ]]
                                   && \text{by the definition of ordering}\\
                                   &= [[ ord varset in iN ]]
       \end{aligned} $

    \item $[[iN = iP → iM]]$\\
       $ \begin{aligned}
        [[ ord varset in [σ]iN ]] &= [[ ord varset in [σ](iP → iM) ]] \\
                                   &= [[ ord varset in ([σ]iP → [σ]iM) ]]
                                   && \text{by the definition of substitution}\\
                                   &= [[ ord varset in [σ]iP, (ord varset in [σ]iM \ {ord varset in [σ]iP}) ]]
                                   && \text{by the definition of ordering}\\
                                   &= [[ ord varset in iP, (ord varset in iM \ {ord varset in iP}) ]]
                                   && \text{by the induction hypothesis}\\
                                   &= [[ ord varset in iP → iM ]]
                                   && \text{by the definition of ordering}\\
                                   &= [[ ord varset in iN ]]
       \end{aligned} $
    \item The proofs of the positive cases are symmetric.
  \end{caseof}
\end{proof}

\begin{lemma}[Completeness of variable ordering]
  \label{lemma:ord-completeness}
  Variable ordering is invariant under equivalence. For arbitrary $[[varset]]$,
   \begin{itemize}
  \item[$-$] If $[[iN ≈ iM]]$ then $[[ord varset in uN]] = [[ord varset in uM]]$ (as lists)
  \item[$+$] If $[[iP ≈ iQ]]$ then $[[ord varset in uP]] = [[ord varset in uQ]]$ (as lists)
  \end{itemize}
\end{lemma}
\begin{proof}
  Mutual induction on $[[iN ≈ iM]]$ and $[[iP ≈ iQ]]$.
\end{proof}

### 2023-08-08 proofs/tex/subst-lemmas.tex
wip: leftovers

\begin{lemma}[Substitution strengthening]
  \label{lemma:subst-restr-fv}
  Restricting the substitution to the free variables of the
  substitution subject does not affect the result.
  Suppose that $[[Γ2 ⊢ σ : Γ1]]$. Then
    \begin{itemize}
  \item[$+$] if $[[Γ1 ⊢ iP]]$ then $[[ [σ]iP ]] = [[ [σ|fv iP]iP ]]$,
  \item[$-$] if $[[Γ1 ⊢ iN]]$ then $[[ [σ]iN ]] = [[ [σ|fv iN]iN ]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  \ilyam{todo}
\end{proof}

>>>>

\begin{lemma}[Substitution strengthening]
  \label{lemma:subst-restr-fv}
  Restricting the substitution to the free variables of the
  substitution subject does not affect the result.
  Suppose that $[[σ]]$ is a substitution,
  $[[iP]]$ and $[[iN]]$ are types. Then
 \begin{itemize}
    \item[$+$] $[[ [σ]iP ]] = [[ [σ|fv iP]iP ]]$,
    \item[$-$] $[[ [σ]iN ]] = [[ [σ|fv iN]iN ]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  First, we strengthen the statement by saying that
  one can restrict the substitution to an arbitrary
  superset of the free variables of the substitution subject:
 \begin{itemize}
    \item[$+$] $[[ [σ]iP ]] = [[ [σ|varset]iP ]]$, for any $[[varset]] \supseteq [[fv iP]]$, and
    \item[$-$] $[[ [σ]iN ]] = [[ [σ|varset]iN ]]$, for any $[[varset]] \supseteq [[fv iN]]$.
  \end{itemize}
  Then the proof is a straightforward induction on $[[iP]]$ and mutually, on $[[iN]]$.
  For the base cases: 
  \begin{caseof}
    \item $[[iN = α⁻]]$\\
      Then $[[ [σ]α⁻ ]] = [[σ|varset]] ([[α⁻]])$ by definition, 
      since $[[α⁻]] \in [[fv α⁻ ⊆ varset]]$.
    \item $[[iN = iP → iM]]$\\
      Then $[[ [σ](iP → iM) ]] = [[ [σ]iP → [σ]iM ]]$ by definition.
      Since $[[fv iP]] \subseteq [[fv(iP → iM)]] \subseteq [[varset]]$, 
      the induction hypothesis is applicable to $[[ [σ]iP ]]$: 
      $[[ [σ]iP ]] = [[ [σ|varset]iP ]]$. 
      Analogously, and $[[ [σ]iM ]] = [[ [σ|varset]iM ]]$.
      Then 
      $[[ [σ](iP → iM) ]] = [[ [σ|varset]iP → [σ|varset]iM ]]
       = [[ [σ|varset](iP → iM) ]]$.
    \item $[[iN = ↑iP]]$ is proved analogously to the previous case.
    \item $[[iN = ∀pas.iM ]]$ (where $[[pas]]$ is not empty)\\
      Then $[[ [σ]∀pas.iM ]] = [[ ∀pas.[σ]iM ]]$ by definition.
      Let us assume $[[pas]]$ are fresh variables, 
      it means that $[[σ]]([[α± ]]) = [[α± ]]$ for any 
      $[[α± ]] \in [[pas]]$, and thus, 
      $[[ [σ|varset] ]] = [[ [σ|(varset ∪ {pas})] ]]$
      immediately from the definition.

      Since $[[varset ⊆ fv(∀pas.iM)]] = [[fv iM \ {pas}]]$,
      $[[varset ∪ {pas} ⊆ fv(iM)]]$.
      Then by the induction hypothesis, 
      $[[ [σ]iM =  [σ|(varset ∪ {pas})]iM ]]$.
      Finally, 
      $[[ [σ]∀pas.iM ]] = [[ ∀pas.[σ|(varset ∪ {pas})]iM ]] = 
      [[ ∀pas.[σ|varset]iM ]] = [[ [σ|varset]∀pas.iM ]]$.
    \item The positive cases are proves symmetrically. 
  \end{caseof}
\end{proof}

### 2023-08-08 proofs/tex/wf-lemmas.tex
wip: leftovers

\begin{lemma}[Well-formedness agrees with substitution]
  \label{lemma:wf-subst}
  Suppose that $[[Γ2 ⊢ σ : Γ1]]$. Then
  \begin{itemize}
  \item[$+$] $[[Γ, Γ1 ⊢ iP]] ~\Leftrightarrow~ [[Γ, Γ2 ⊢ [σ]iP]]$
  \item[$-$] $[[Γ, Γ1 ⊢ iN]] ~\Leftrightarrow~ [[Γ, Γ2 ⊢ [σ]iN]]$
  \end{itemize}
\end{lemma}
\begin{proof}
  \ilyam{todo}
\end{proof}

\begin{corollary}
  \label{lemma:wf-subst}
  Suppose that $[[Γ2 ⊢ σ : Γ1]]$. Then
  \begin{itemize}
  \item[$+$] $[[Γ1, Γ2 ⊢ iP]] ~\Leftrightarrow~ [[Γ2 ⊢ [σ]iP]]$
  \item[$-$] $[[Γ1, Γ2 ⊢ iN]] ~\Leftrightarrow~ [[Γ2 ⊢ [σ]iN]]$
  \end{itemize}
\end{corollary}

\begin{lemma}[Equivalent Contexts]
  \label{lemma:wf-ctxt-equiv}
  In the well-formedness judgment, only used variables matter:
  \begin{itemize}
  \item[$+$] if $[[{Γ1} ∩ fv iP]] = [[{Γ2} ∩ fv iP]]$ then
    $[[Γ1 ⊢ iP]] \iff [[Γ2 ⊢ iP]]$,
  \item[$-$] if $[[{Γ1} ∩ fv iN]] = [[{Γ2} ∩ fv iN]]$ then
    $[[Γ1 ⊢ iN]] \iff [[Γ2 ⊢ iN]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  By simple mutual induction on $[[iP]]$ and $[[iQ]]$. 
\end{proof}

@@

>>>>

\begin{lemma}[Equivalent Contexts]
  \label{lemma:wf-ctxt-equiv}

  In the well-formedness judgment, only used variables matter:
  \begin{itemize}
  \item[$+$] if $[[{Γ1} ∩ fv iP]] = [[{Γ2} ∩ fv iP]]$ then
    $[[Γ1 ⊢ iP]] \iff [[Γ2 ⊢ iP]]$,
  \item[$-$] if $[[{Γ1} ∩ fv iN]] = [[{Γ2} ∩ fv iN]]$ then
    $[[Γ1 ⊢ iN]] \iff [[Γ2 ⊢ iN]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  By simple mutual induction on $[[iP]]$ and $[[iQ]]$. 
\end{proof}

\begin{lemma}[Well-formedness Context Weakening]
  \label{lemma:wf-weakening}
  Suppose that $[[{Γ1} ⊆ {Γ2}]]$, then
  \begin{itemize}
    \item[$+$] if $[[Γ1 ⊢ iP]]$ then $[[Γ2 ⊢ iP]]$,
    \item[$-$] if $[[Γ1 ⊢ iN]]$ then $[[Γ2 ⊢ iN]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  By simple mutual induction on $[[iP]]$ and $[[iQ]]$. 
\end{proof}

@@

\begin{lemma}[Well-formedness agrees with substitution]
  \label{lemma:wf-subst}
  Suppose that $[[Γ2 ⊢ σ : Γ1]]$. Then
  \begin{itemize}
  \item[$+$] $[[Γ, Γ1 ⊢ iP]]$ implies $[[Γ, Γ2 ⊢ [σ]iP]]$, and
  \item[$-$] $[[Γ, Γ1 ⊢ iN]]$ implies $[[Γ, Γ2 ⊢ [σ]iN]]$.
  \end{itemize}
\end{lemma}
\begin{proof}
  We prove it by induction on $[[Γ, Γ1 ⊢ iP]]$ and mutually, on $[[Γ, Γ1 ⊢ iN]]$.
  Let us consider the last rule used in the derivation.
  \begin{caseof}
    \item \ruleref{\ottdruleWFTPVarLabel}, 
      i.e. $[[iP]]$ is $[[α⁺]]$.\\
      By inversion, $[[α⁺ ∊ {Γ, Γ1}]]$, then
      \begin{itemize}
        \item if $[[α⁺ ∊ {Γ1}]]$ then $[[ Γ2 ⊢ [σ]α⁺ ]]$, 
          and by weakening (\cref{lemma:wf-weakening}),
          $[[ Γ, Γ2 ⊢ [σ]α⁺ ]]$;
        \item if $[[α⁺ ∊ {Γ} \ {Γ1}]]$ then $[[ [σ]α⁺ = α⁺ ]]$,
          and by \ruleref{\ottdruleWFTPVarLabel}, $[[ Γ, Γ2 ⊢ α⁺ ]]$.
      \end{itemize}

    \item \ruleref{\ottdruleWFTShiftULabel},
      i.e. $[[iP]]$ is $[[↓iN]]$.\\
      Then $[[Γ, Γ1 ⊢ ↓iN]]$ means $[[Γ, Γ1 ⊢ iN]]$ by inversion,
      and by the induction hypothesis, $[[Γ, Γ2 ⊢ [σ]iN]]$.
      Then by  \ruleref{\ottdruleWFTShiftULabel}, $[[Γ, Γ2 ⊢ ↓[σ]iN]]$, 
      which by definition of substitution is rewritten as $[[Γ, Γ2 ⊢ [σ]↓iN]]$.

    \item \ruleref{\ottdruleWFTExistsLabel},
      i.e. $[[iP]]$ is $[[∃nas.iQ]]$.\\
      Then $[[Γ, Γ1 ⊢ ∃nas.iQ]]$ means $[[Γ, nas, Γ1 ⊢ iQ]]$ 
      by inversion, and by the induction hypothesis, 
      $[[Γ, nas, Γ2 ⊢ [σ]iQ]]$.
      Then by  \ruleref{\ottdruleWFTExistsLabel}, 
      $[[Γ, nas, Γ2 ⊢ ∃nas.[σ]iQ]]$, 
      which by definition of substitution is rewritten as 
      $[[Γ, Γ2 ⊢ [σ]∃nas.iQ]]$.

    \item The negative cases are proved symmetrically.
  \end{caseof}

\end{proof}
